\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-23 paper.tex
Gregor comments
standardise Big and Little ARG

% TODO: Provide a map of the paper's sections and how they tie together in support of gARG as a way forward.
% Here is a rough draft - it would be good to state the most important message (in one sentence or so) from each of the sections to help with frontloading key messages
% Genome ARGs &
% Event ARGs
In the following, we first define the genome ARG encoding
and contrast it with the event ARG approach.
We then discuss the important concept of ancestral material in
Section~\ref{sec-ancestral-material}
and [DO SOME STUFF].
% Converting and eARG to gARG &
% ARGs and local trees &
% Locally unary nodes &
% Levels of simplification
This simplification is closely related to operations needed to convert eARGs to
various forms of gARGs,
which opens a rich set of details about ARGs and local trees, local unary nodes, and
levels of simplification.
% Precision of recombination information
We then highlight the important difference between the required level of precision
about recombination information for eARGs compared to gARGs.
% Computational efficiency and data interchange
The way this recombination information is stored in eARGs and gARGs has important
implications for computational efficiency, which is critical if we want to use ARGs
for modern genomic datasets.
% Discussion
We wrap with a discussion highlighting that WHAT-IS-THE-KEY-SINGLE-MESSAGE-NOT-COVERED-ABOVE?
Finally, the literature on ARGs and associated aspects is vast, particularly
due to different definitions and implementations. We cover some important aspects
in appendices, including
a brief history of ancestral graphs (Appendix~\ref{sec-arg-history}),
description of the big and little ARGs (Appendix~\ref{sec-big-and-little-arg}),
survey of ARG inference methods (Appendix~\ref{sec-survey-arg-infer}), and
ARGs at an individual vs cell lineage level (Appendix~\ref{sec-cell-lineages-and-args}).

@@

In many settings, nodes are dated, i.e.\ each
node $u\in N$ is associated with a time $\tau_u$.
The topological relationships
defined by the directed graph structure, however,
 is sufficient for many applications,
and we will assume that node dates are not known here.
% This should be obvious, but seems like it needs to be said
In practical settings, we will wish to associate additional
metadata with nodes such as sample identifiers or quality-control metrics;
It is therefore best to think of the
integers used here in the definition of a node as an \emph{identifier},
with which arbitrary additional information can be associated.
Concretely defining nodes as integers in this way has some
significant advantages. In particular, it provides a direct link
between an ARG and its local trees in terms of a well-defined
combinatorial object (Section~\ref{sec-ARG-and-local-trees}),
as well as enabling straightforward connection between
definitions and practical implementation (Section~\ref{sec-efficiency}).

@@

\section{Event ARGs}
\label{sec-eARG}
% % I feel like sophisticated readers would at this point be saying,
% % ah, yeah, that's basically what an ARG is just in different words;
% % what's the point of this paper?
The genome ARG framework defined in the previous section is
in many ways a clarification of existing treatments
\cite[e.g.][]{mathieson2020ancestry,shipilina2023origin},
adding concrete details to describe the
differential inheritance of genetic material between genomes.
% This is definitely needed for the Nicks, who are always thinking
% about the deeper processes and are quite happy being loosey-goosey
% with representation details
Note that here, and throughout, we are not questioning the form of the
actual ancestral processes that occur in nature, but rather how we
\emph{represent} the outcomse of such processes in a practical sense.
% Again, need to remind the reader that there's an actual point
% to all this getting stuck into the details.
These practical details,
as demonstrated through the remainder of the article, have important
consequences not only for how methods exchange information about
simulated and inferred ARGs, but more fundamentally in
how we set our goals for inference and evaluate the success of results.

@@

\section{Ancestral material and sample resolution}
\label{sec-ancestral-material}
Ancestral material~\citep{wiuf1999ancestry,wiuf1999recombination},
denoting the genomic intervals ancestral to a set of samples
on the edges of an ARG, is a key concept. In
Fig.~\ref{fig-arg-in-pedigree}, for example, we have
four sample genomes, \noderef{a}--\noderef{d}. As we
trace their genetic ancestry into the previous generation
(\noderef{e}--\noderef{h}), we can think of the ancestral
material of the sample propagating through the graph
backwards in time. In the interval $[2, 7)$, there is a
local coalescence where nodes \noderef{a} and \noderef{c}
find common ancestor in \noderef{f}. Thus, in this interval,
the total number of genome segments that are ancestral to the
sample is reduced from four to three. Note that in the same
interval, node \noderef{e} is not ancestral to any
of the samples: it carries \emph{non ancestral} material.
This process of local coalescence continues through the
graph, until all samples reach their most recent common
ancestor in node \noderef{n}.

The process of tracking local coalescences and updating
segments of ancestral material is a core element of
Hudson's seminal simulation
algorithm~\citep{hudson1983testing,kelleher2016efficient},
and the key distinguishing feature between the
``big'' and ``little'' ARG stochastic processes
(see Appendix~\ref{sec-big-and-little-arg}).
The ability to \emph{store} the resolved ancestral material
for a sample
in an ARG is also a key distinction between the eARG and gARG
encodings. Because an eARG stores only the graph topology and
recombination breakpoints, there is no way to locally
ascertain the ancestral material without traversing the graph
rootwards from the samples,
resolving the effects of recombination and common ancestor events.

@@

The sample-resolved gARG of Fig.~\ref{fig-ancestry-resolution}C
differs in some important ways to the
original eARG (Fig.~\ref{fig-ancestry-resolution}A).
Firstly, we can see that some nodes and edges have been removed entirely
from the graph.
The ``grand MRCA'' \noderef{q} is normally omitted from the
sample-resolved gARG because all segments of the genome have
fully coalesced before it is reached. Likewise, the edge
between \noderef{g} and \noderef{j} is omitted because the recombination
event at position $5$ (represented by node \noderef{g})
fell in non-ancestral material.
More generally, we can see that the sample resolved
gARG of Fig.~\ref{fig-ancestry-resolution}C
allows for ``local'' inspection
of an ARG in a way that is not possible by storing
common ancestor and recombination events in an eARG. Because
the ancestral material is stored with each edge, the
cumulative effects of events over time can be reasoned
about, without first ``replaying'' those events. Many computations
that we wish to perform on an ARG will require resolving
the ancestral material with respect to a sample.
% this is weak, but would be good to emphasise this point.
% you just end up running simplify lots of times, in your
% downstream algs
The gARG encoding
allows us to perform this resolution of ancestral material once
and to store the result,
whereas the eARG encoding requires us to repeat the process
each time.

@@

The scalability gains made by recent ARG inference methods have been, in part, due to inferring
lower levels of precision about recombination. Indeed, a common statement in the literature is
that modern tools such as \tsinfer~\citep{kelleher2019inferring} and \relate~\citep{speidel2019method}
construct not an ARG but ``only'' a sequence of local
trees \citep{hejase2020summary, ignatieva2021kwarg}. While this statement
is correct inasmuch as neither method attempts to infer the explicit recombination events that cause
adjacent trees to differ, it is certainly not the case that they infer independent local trees.
In fact, both methods take advantage of genome-wide mutation information to construct genealogies,
and as a result nodes and edges persist across multiple local trees.
Their results can therefore be thought of as an inheritance graph, and represented in the gARG format.

@@

Note that both \kwarg\ and \argweaver\ produce many ARGs
(those shown in Fig.~\ref{fig-inferred-args}A,B are arbitrarily selected examples).
This stems from the fact that \kwarg\ searches the space of plausible ARGs,
outputting minimal ones using heuristics, and \argweaver\ samples
from the posterior distribution over
ARGs given the data and a fixed model (the SMC$^\prime$ \citep{marjoram2006fast},
an approximation of the full coalescent with recombination).
In contrast, \tsinfer\ and \relate\ (Fig.~\ref{fig-inferred-args}C,D) are
deterministic methods, aiming to find a single ``best guess'' ARG.
The ability of a gARG to encode the effects of recombination, rather
than having to precisely specify each event, is a important contributor to this aim,
as most recombination events cannot be uniquely identified from the data.
Indeed, there is considerable variation in the number of recombinations inferred by
different methods, with e.g.~\argweaver\ (Fig.~\ref{fig-inferred-args} B) suggesting
far more than are required under minimal parsimony assumptions (Fig.~\ref{fig-inferred-args} A).

% Here we describe recombination features
A rough feel for the amount of recombination in each ARG is given by the distribution
of number of parents per node, as highlighted by the node colour scheme.
In Fig.~\ref{fig-inferred-args}A,B, each recombination event is tagged by
a node with exactly 2 parents and one child. Assuming that the breakpoint for each event
occurs at a different location (the "infinite breakpoints" model), then the
number of breakpoints should exactly equal the number of 2-parent (brown) nodes in the plot.
In contrast, Figure Fig.~\ref{fig-inferred-args}C,D have ARG nodes with multiple
parents and multiple children. These indicate deliberate imprecision, where one
or more recombinations have been ``pushed down'' onto a more recent, identifiable node.
Several of these nodes have more than two parents (in red), representing
multiple recombinations inferred to have occurred in an unresolvable order.
Nevertheless, the number of breakpoints in these two ARGs is smaller than the
number of multiple-parent ARG nodes, showing that several multiple-parent nodes
must share breakpoint positions. The sharing of breakpoint locations
between these nodes indicates that in \tsinfer\ and \relate\ ARGs it may be
difficult to condense each transition between trees into a set of simple SPR moves.
Overall, Fig.~\ref{fig-inferred-args} shows that \tsinfer\ and \relate\ ARGs contain a level
of detail about recombination that lies somewhere between a sequence of unrelated local
trees on one extreme and an ARG with precisely specified recombination events on the other.

@@

% FIXME this needs more work and doesn't really fit into the
% narrative right now, but we do need to make this point.
The key insight that makes the succinct tree sequence encoding
an efficient substrate for defining analysis algorithms is that
it allows us to generate the local trees along the genome
in a way that allows us to reason about the \emph{differences}
between those trees.
Sequentially generating the local trees along the genome
is also fundamental, and is necessary whenever we need to
perform calculations that are contingent on more than just the
isolated properties of an edge. \cite{kelleher2016efficient}
showed how all trees can be sequentially generated in
constant time per tree transition in a fully simplified gARG.
Furthermore, we can easily reason about how tree topologies
change (and stay the same), leading to efficient algorithms
for computing population genetic
statistics~\citep{kelleher2016efficient,ralph2020efficiently},
implementing the Li and Stephens
model~\citep{kelleher2019inferring,wohns2022unified}
and likely many more.

@@

\section{The Big and Little ARG}
\label{sec-big-and-little-arg}
Here we review the two predominant stochastic processes which construct ARGs:
the ``big" ARG process of \cite{griffiths1997ancestral}, and the ``little" ARG process of
 \cite{hudson1983properties}. The big ARG process is mathematically simpler
 but is computationally intractable due to generating a vast number of ancestors
 which contribute no genetic material to the initial sample.
The little ARG process avoids non-genetic ancestors at the cost of more complex
dynamics and state space. We also demonstrate that evaluating the sampling probability
of either process---a key quantity in many statistical approaches---requires that the
gARG (or eARG) data structure be interpreted in a model-specific way.

A generic state of the little ARG process consists of a finite collection of lineages $L$,
each of which is a list of disjoint ancestry segments $(\ell, r, a)$, where
$[\ell, r)$ is a half-closed genomic interval and $a$ is an integer
tracking the number of samples to which the lineage is ancestral over that interval.
We also usually track the node associated with each segment, but
that is not important for our purposes here so we omit it to lighten notation.
The initial condition for a sample of $n$ genomes of length $m$ consists of $n$ lineages
of the form $\{(0, m, 1)\}$. The process traverses a series of common ancestor and
recombination events backwards in time.
Recombination events happen at rate $\rho \nu / (m - 1)$,
where $\rho \geq 0$ is a per-genome recombination rate and
 \[
 \nu = \sum_{x \in L}\left( \max_{(\ell, r, a) \in x}r
     - \min_{(\ell, r, a) \in x}\ell - 1 \right)
 \]
 is the number of available ``links" surrounded by ancestral material.
 At a recombination event we choose one of these links uniformly and break it,
 replacing the original lineage in $L$ with two new lineages containing the ancestral material
 to the left and right of the break point, respectively.

@@

In the Griffiths formulation (the big ARG process), each edge in the graph corresponds to an extant
lineage and nodes are events in the process. The $n$ initial leaf nodes are
sampling events. Common ancestor events occur at rate $\binom{|L|}{2}$.
When a common ancestor event happens, two uniformly chosen lineages
merge into a common ancestor lineage.
Recombination events happen at rate $|L| \rho$. Here, we choose a lineage (i.e.\ edge) uniformly,
and a breakpoint $0 < x < m$ uniformly on its genome. We terminate the edge at a
node, record the breakpoint, and start two new edges from this node. The process
then continues until there is only one lineage left (the Grand Most Recent
Common Ancestor, GMRCA), which is guaranteed to
happen in finite time because of the quadratic rate of coalescing vs.\ linear rate of branching.

The state-space of the big ARG process is much simpler than that of the little ARG process,
which greatly facilitates mathematical reasoning. This simplicity comes at a
substantial cost, however, if we wish to use it as a practical means of
simulating recombinant ancestries.
The number of events in the big ARG all the way back to the GMRCA
is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
of events required to simulate the little ARG is
$O(\rho^2)$~\citep{hein2004gene,baumdicker2021efficient}.
This disparity arises because the majority of the events in the big ARG are
recombination events which occur outside of ancestral material,
and this do not have any bearing on the ancestry of the initial sample.
Because we don't keep track of the distribution of ancestral material during the process,
we generate a vastly larger graph.

Figure \ref{hudson_vs_bigARG} illustrates the more complex state space
of the little ARG process, as well as the extra events which occur in the big ARG process.
Moreover, it depicts the rates of common ancestors and recombination events in each
interval of time of the realisations.
In order to evaluate these rates, and hence the sampling probability
\cite[see e.g.][Equation (3)]{mahmoudi2022bayesian},
it is necessary to know the number of lineages and number of extant links
available for recombination in each time interval.
Some representations may not provide this information.
For example, in the gARG encoding depicted in Figure \ref{fig-ancestry-resolution}C
it is clear that a recombination takes place between nodes \noderef{b} and
\noderef{d} as well as \noderef{e}, but the exact time of the event is ambiguous,
and thus so is the number of lineages during the time interval. In fact, this
information cannot be recovered from the gARG encoding used in Figure \ref{fig-ancestry-resolution}C,
and alternative conventions must be introduced to resolve such ambiguities.
For the likelihood-based inference algorithms for the coalescent with recombination,
it is sufficient to create \emph{two} gARG nodes at the time of the recombination event:
see the appendix of \cite{baumdicker2021efficient} for details of evaluating
coalescent with recombination likelihoods using this convention.
This is also the interpretation depicted in
Figure \ref{hudson_vs_bigARG}, but it means that the two edges above node \noderef{b}
in Figure \ref{fig-ancestry-resolution} should correspond to only one lineage,
along which all $m-1$ links are available for recombination.
The lineage then splits into two at the time of nodes \noderef{d} and \noderef{e}.
Nodes \noderef{k}, \noderef{l}, and \noderef{m} in Figure \ref{fig-ancestry-resolution}
demonstrate that the same issue can affect the number of links available for recombination:
without an external convention, the exact time at which the trapped ancestral material on
node \noderef{k} ceases to be available for an effective recombination in the little ARG process.

@@

	% Gray dashed lines to visually separate holding times
	\draw[color=gray, dashed] (7.5, 1.2) -- (9.1, 1.2);
	\draw[color=gray, dashed] (7.5, 1.7) -- (9.1, 1.7);
	\draw[color=gray, dashed] (7.5, 2.4) -- (9.1, 2.4);
	\draw[color=gray, dashed] (7.5, 3.1) -- (9.1, 3.1);
	\draw[color=gray, dashed] (7.5, 3.8) -- (9.1, 3.8);
	\draw[color=gray, dashed] (7.5, 4.5) -- (9.1, 4.5);
	\draw[color=gray, dashed] (7.5, 5.4) -- (9.1, 5.4);
	\draw[color=gray, dashed] (7.5, 6) -- (9.1, 6);
	\draw[color=gray, dashed] (7.5, 6.6) -- (9.1, 6.6);
	\draw[color=gray, dashed] (7.5, 7.2) -- (9.1, 7.2);
\end{tikzpicture}
\end{tabular}
}
\caption{(A)
A realisation of the graph traversed by Hudson's algorithm started from a
sample of three chromosomes with $m$ discrete sites each at time $t = 0$, and
propagated until time $T$. The MRCA on the genetic interval $[v, w)$ is reached
at event \noderef{b}, while that on $[0, v)$ is reached at event \noderef{c}.
The non-ancestral segment $[v, w)$ above
\noderef{a} contributes to the rate of effective recombinations because it
is trapped between ancestral segments. The two columns titled CA and RE
are the respective rates of mergers and recombinations when
the recombination rate is $\rho$.
(B) A corresponding realisation of a big ARG, which augments Hudson's algorithm
by tracking nonancestral lineages. The result is a simpler state space and
dynamics, at the cost of extra nodes and edges, highlighted in red, which do
not affect the local tree at any site.
Recombination positions are labelled alphabetically in time, and their ordering
along the genome is $y < v < x < w$, of which the first only appears in panel B.
There are two separate recombination events at link $w$.
}
\label{hudson_vs_bigARG}
\end{figure}

>>>>

% TODO: Provide a map of the paper's sections and how they tie together in support of gARG as a way forward.
% Here is a rough draft - it would be good to state the most important message (in one sentence or so) from each of the sections to help with frontloading key messages
% Genome ARGs &
% Event ARGs
In the following, we first define the genome ARG encoding
and contrast it with the event ARG approach.
We then discuss the important concept of ancestral material in
Section~\ref{sec-ancestral-material}
and [DO SOME STUFF].
% Converting and eARG to gARG &
% ARGs and local trees &
% Locally unary nodes &
% Levels of simplification
This simplification is closely related to operations needed to convert eARGs to
various forms of gARGs,
which opens a rich set of details about ARGs and local trees, local unary nodes, and
levels of simplification.
% Precision of recombination information
We then highlight the important difference between the required level of precision
about recombination information for eARGs compared to gARGs.
% Computational efficiency and data interchange
The way this recombination information is stored in eARGs and gARGs has important
implications for computational efficiency, which is critical if we want to use ARGs
for modern genomic datasets.
% Discussion
We wrap with a discussion highlighting that WHAT-IS-THE-KEY-SINGLE-MESSAGE-NOT-COVERED-ABOVE?
Finally, the literature on ARGs and associated aspects is vast, particularly
due to different definitions and implementations. We cover some important aspects
in appendices, including
a brief history of ancestral graphs (Appendix~\ref{sec-arg-history}),
description of the Big and Little ARGs (Appendix~\ref{sec-big-and-little-arg}),
survey of ARG inference methods (Appendix~\ref{sec-survey-arg-infer}), and
ARGs at an individual vs cell lineage level (Appendix~\ref{sec-cell-lineages-and-args}).

@@

In many settings, nodes are dated, i.e.\ each
node $u\in N$ is associated with a time $\tau_u$.
The topological relationships
defined by the directed graph structure, however,
 is sufficient for many applications,
and we will assume that node dates are not known here.
% This should be obvious, but seems like it needs to be said
In practical settings, we will wish to associate additional
metadata with nodes such as sample identifiers or quality-control metrics.
It is therefore best to think of the
integers used here in the definition of a node as an \emph{identifier},
with which arbitrary additional information can be associated.
Concretely defining nodes as integers in this way has some
significant advantages. In particular, it provides a direct link
between an ARG and its local trees in terms of a well-defined
combinatorial object (Section~\ref{sec-ARG-and-local-trees}),
as well as enabling straightforward connection between
definitions and practical implementation (Section~\ref{sec-efficiency}).

@@

\section{Event ARGs}
\label{sec-eARG}
% % I feel like sophisticated readers would at this point be saying,
% % ah, yeah, that's basically what an ARG is just in different words;
% % what's the point of this paper?
The genome ARG framework defined in the previous section is
in many ways a clarification of existing treatments
\cite[e.g.][]{mathieson2020ancestry,shipilina2023origin},
adding concrete details to describe the
differential inheritance of genetic material between genomes.
% This is definitely needed for the Nicks, who are always thinking
% about the deeper processes and are quite happy being loosey-goosey
% with representation details
Note that here, and throughout, we are not questioning the form of the
actual ancestral processes that occur in nature, but rather how we
\emph{represent} the outcomes of such processes in a practical sense.
% Again, need to remind the reader that there's an actual point
% to all this getting stuck into the details.
These practical details,
as demonstrated through the remainder of the article, have important
consequences not only for how methods exchange information about
simulated and inferred ARGs, but more fundamentally in
how we set our goals for inference and evaluate the success of results.

@@

\section{Ancestral material and sample resolution}
\label{sec-ancestral-material}
Ancestral material~\citep{wiuf1999ancestry,wiuf1999recombination}
is a key concept in understanding the overall inheritance structure
of an ARG. It denotes the genomic intervals ancestral to a set of samples
on the edges of an ARG.
In Fig.~\ref{fig-arg-in-pedigree}, for example, we have
four sample genomes, \noderef{a}--\noderef{d}. As we
trace their genetic ancestry into the previous generation
(\noderef{e}--\noderef{h}), we can think of the ancestral
material of the samples propagating through the graph
backwards in time. In the interval $[2, 7)$, there is a
local coalescence where nodes \noderef{a} and \noderef{c}
find a common ancestor in \noderef{f}. Thus, in this interval,
the total number of genome segments that are ancestral to the
sample is reduced from four to three. Note that in the same
interval, node \noderef{e} is not ancestral to any
of the samples: it carries \emph{non ancestral} material.
This process of local coalescence continues through the
graph, until all samples reach their most recent common
ancestor in node \noderef{n}.

The process of tracking local coalescences and updating
segments of ancestral material is a core element of
Hudson's seminal simulation simulation
algorithm~\citep{hudson1983testing,kelleher2016efficient},
and the key distinguishing feature between the
``Big'' and ``Little'' ARG stochastic processes
(see Appendix~\ref{sec-big-and-little-arg}).
The ability to \emph{store} the resolved ancestral material
for a sample
in an ARG is also a key distinction between the eARG and gARG
encodings. Because an eARG stores only the graph topology and
recombination breakpoints, there is no way to locally
ascertain the ancestral material without traversing the graph
rootwards from the samples,
resolving the effects of recombination and common ancestor events.

@@

The sample-resolved gARG of Fig.~\ref{fig-ancestry-resolution}C
differs in some important ways to the
original eARG (Fig.~\ref{fig-ancestry-resolution}A).
Firstly, we can see that some nodes and edges have been removed entirely
from the graph.
The ``grand MRCA'' \noderef{q} is omitted from the
sample-resolved gARG because all segments of the genome have
fully coalesced before it is reached. Likewise, the edge
between \noderef{g} and \noderef{j} is omitted because the recombination
event at position $5$ (represented by node \noderef{g})
fell in non-ancestral material.
More generally, we can see that the sample resolved
gARG of Fig.~\ref{fig-ancestry-resolution}C
allows for ``local'' inspection
of an ARG in a way that is not possible by storing
common ancestor and recombination events in an eARG. Because
the ancestral material is stored with each edge, the
cumulative effects of events over time can be reasoned
about, without first ``replaying'' those events. Many computations
that we wish to perform on an ARG will require resolving
the ancestral material with respect to a sample.
% this is weak, but would be good to emphasise this point.
% you just end up running simplify lots of times, in your
% downstream algs
The gARG encoding
allows us to perform this resolution of ancestral material once
and to store the result,
whereas the eARG encoding requires us to repeat the process
each time.

@@

The scalability gains made by recent ARG inference methods have been, in part, due to inferring
lower levels of precision about recombination. Indeed, a common statement in the literature is
that modern tools such as \tsinfer~\citep{kelleher2019inferring} and \relate~\citep{speidel2019method}
construct not an ARG but ``only'' a sequence of local
trees \citep[e.g.][]{hejase2020summary, brandt2021evaluation, ignatieva2021kwarg}. While this statement
is correct inasmuch as neither method attempts to infer the explicit recombination events that cause
adjacent trees to differ, it is certainly not the case that they infer independent local trees.
In fact, both methods take advantage of genome-wide mutation information to construct genealogies,
and as a result nodes and edges persist across multiple local trees.
Their results can therefore be thought of as an inheritance graph, and represented in the gARG format.

@@

Note that both \kwarg\ and \argweaver\ produce many ARGs
(those shown in Fig.~\ref{fig-inferred-args}A,B are arbitrarily selected examples).
This stems from the fact that \kwarg\ searches the space of plausible ARGs,
outputting minimal ones using heuristics, and \argweaver\ samples
from the posterior distribution over
ARGs given the data and a fixed model (a discretised version of the SMC$^\prime$ \citep{marjoram2006fast},
an approximation of the full coalescent with recombination).
In contrast, \tsinfer\ and \relate\ (Fig.~\ref{fig-inferred-args}C,D) are
deterministic methods, aiming to find a single ``best guess'' ARG.
The ability of a gARG to encode the effects of recombination, rather
than having to precisely specify each event, is a important contributor to this aim,
as most recombination events cannot be uniquely identified from the data.
Indeed, there is considerable variation in the number of recombinations inferred by
different methods, with e.g.~\argweaver\ (Fig.~\ref{fig-inferred-args} B) suggesting
far more than are required under minimal parsimony assumptions (Fig.~\ref{fig-inferred-args} A).

% Here we describe recombination features
A rough feel for the amount of recombination in each ARG is given by the distribution
of number of parents per node, as highlighted by the node colour scheme.
In Fig.~\ref{fig-inferred-args}A,B, each recombination event is tagged by
a node with exactly 2 parents and one child. Assuming that the breakpoint for each event
occurs at a different location (the ``infinite breakpoints" model), then the
number of breakpoints should exactly equal the number of 2-parent (brown) nodes in the plot.
In contrast, Figure Fig.~\ref{fig-inferred-args}C,D have ARG nodes with multiple
parents and multiple children. These indicate deliberate imprecision, where one
or more recombinations have been ``pushed down'' onto a more recent, identifiable node.
Several of these nodes have more than two parents (in red), representing
multiple recombinations inferred to have occurred in an unresolvable order.
Nevertheless, the number of breakpoints in these two ARGs is smaller than the
number of multiple-parent ARG nodes, showing that several multiple-parent nodes
must share breakpoint positions. The sharing of breakpoint locations
between these nodes indicates that in \tsinfer\ and \relate\ ARGs it may be
difficult to condense each transition between trees into a set of simple SPR moves.
Overall, Fig.~\ref{fig-inferred-args} shows that \tsinfer\ and \relate\ ARGs contain a level
of detail about recombination that lies somewhere between a sequence of unrelated local
trees on one extreme and an ARG with precisely specified recombination events on the other.

@@

% FIXME this needs more work and doesn't really fit into the
% narrative right now, but we do need to make this point.
The key insight that makes the succinct tree sequence encoding
an efficient substrate for defining analysis algorithms is that
it allows us to recover the local trees along the genome
in a way that allows us to reason about the \emph{differences}
between those trees.
Sequentially generating the local trees along the genome
is also fundamental, and is necessary whenever we need to
perform calculations that are contingent on more than just the
isolated properties of an edge. \cite{kelleher2016efficient}
showed how all trees can be sequentially generated in
constant time per tree transition in a fully simplified gARG.
Furthermore, we can easily reason about how tree topologies
change (and stay the same), leading to efficient algorithms
for computing population genetic
statistics~\citep{kelleher2016efficient,ralph2020efficiently},
implementing the Li and Stephens
model~\citep{kelleher2019inferring,wohns2022unified}
and likely many more.

@@

\section{The Big and Little ARG}
\label{sec-big-and-little-arg}
Here we review the two predominant stochastic processes which construct ARGs:
the ``big" ARG process of \cite{griffiths1997ancestral}, and the ``little" ARG process of
 \cite{hudson1983properties}. The Big ARG process is mathematically simpler
 but is computationally intractable due to generating a vast number of ancestors
 which contribute no genetic material to the initial sample.
The Little ARG process avoids non-genetic ancestors at the cost of more complex
dynamics and state space. We also demonstrate that evaluating the sampling probability
of either process---a key quantity in many statistical approaches---requires that the
gARG (or eARG) data structure be interpreted in a model-specific way.

A generic state of the Little ARG process consists of a finite collection of lineages $L$,
each of which is a list of disjoint ancestry segments $(\ell, r, a)$, where
$[\ell, r)$ is a half-closed genomic interval and $a$ is an integer
tracking the number of samples to which the lineage is ancestral over that interval.
We also usually track the node associated with each segment, but
that is not important for our purposes here so we omit it to lighten notation.
The initial condition for a sample of $n$ genomes of length $m$ consists of $n$ lineages
of the form $\{(0, m, 1)\}$. The process traverses a series of common ancestor and
recombination events backwards in time.
Recombination events happen at rate $\rho \nu / (m - 1)$,
where $\rho \geq 0$ is a per-genome recombination rate and
 \[
 \nu = \sum_{x \in L}\left( \max_{(\ell, r, a) \in x}r
     - \min_{(\ell, r, a) \in x}\ell - 1 \right)
 \]
 is the number of available ``links" surrounded by ancestral material.
 At a recombination event we choose one of these links uniformly and break it,
 replacing the original lineage in $L$ with two new lineages containing the ancestral material
 to the left and right of the break point, respectively.

@@

In the Griffiths formulation (the Big ARG process), each edge in the graph corresponds to an extant
lineage and nodes are events in the process. The $n$ initial leaf nodes are
sampling events. Common ancestor events occur at rate $\binom{|L|}{2}$.
When a common ancestor event happens, two uniformly chosen lineages
merge into a common ancestor lineage.
Recombination events happen at rate $|L| \rho$. Here, we choose a lineage (i.e.\ edge) uniformly,
and a breakpoint $0 < x < m$ uniformly on its genome. We terminate the edge at a
node, record the breakpoint, and start two new edges from this node. The process
then continues until there is only one lineage left (the Grand Most Recent
Common Ancestor, GMRCA), which is guaranteed to
happen in finite time because of the quadratic rate of coalescing vs.\ linear rate of branching.

The state-space of the Big ARG process is much simpler than that of the Little ARG process,
which greatly facilitates mathematical reasoning. This simplicity comes at a
substantial cost, however, if we wish to use it as a practical means of
simulating recombinant ancestries.
The number of events in the Big ARG all the way back to the GMRCA
is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
of events required to simulate the Little ARG is
$O(\rho^2)$~\citep{hein2004gene,baumdicker2021efficient}.
This disparity arises because the majority of the events in the Big ARG are
recombination events which occur outside of ancestral material,
and this do not have any bearing on the ancestry of the initial sample.
Because we don't keep track of the distribution of ancestral material during the process,
we generate a vastly larger graph.

Figure \ref{hudson_vs_bigARG} illustrates the more complex state space
of the Little ARG process, as well as the extra events which occur in the Big ARG process.
Moreover, it depicts the rates of common ancestors and recombination events in each
interval of time of the realisations.
In order to evaluate these rates, and hence the sampling probability
\cite[see e.g.][Equation (3)]{mahmoudi2022bayesian},
it is necessary to know the number of lineages and number of extant links
available for recombination in each time interval.
Some representations may not provide this information.
For example, in the gARG encoding depicted in Figure \ref{fig-ancestry-resolution}C
it is clear that a recombination takes place between nodes \noderef{b} and
\noderef{d} as well as \noderef{e}, but the exact time of the event is ambiguous,
and thus so is the number of lineages during the time interval. In fact, this
information cannot be recovered from the gARG encoding used in Figure \ref{fig-ancestry-resolution}C,
and alternative conventions must be introduced to resolve such ambiguities.
For the likelihood-based inference algorithms for the coalescent with recombination,
it is sufficient to create \emph{two} gARG nodes at the time of the recombination event:
see the appendix of \cite{baumdicker2021efficient} for details of evaluating
coalescent with recombination likelihoods using this convention.
This is also the interpretation depicted in
Figure \ref{hudson_vs_bigARG}, but it means that the two edges above node \noderef{b}
in Figure \ref{fig-ancestry-resolution} should correspond to only one lineage,
along which all $m-1$ links are available for recombination.
The lineage then splits into two at the time of nodes \noderef{d} and \noderef{e}.
Nodes \noderef{k}, \noderef{l}, and \noderef{m} in Figure \ref{fig-ancestry-resolution}
demonstrate that the same issue can affect the number of links available for recombination:
without an external convention, the exact time at which the trapped ancestral material on
node \noderef{k} ceases to be available for an effective recombination in the Little ARG process.

@@

	% Gray dashed lines to visually separate holding times
	\draw[color=gray, dashed] (7.5, 1.2) -- (9.1, 1.2);
	\draw[color=gray, dashed] (7.5, 1.7) -- (9.1, 1.7);
	\draw[color=gray, dashed] (7.5, 2.4) -- (9.1, 2.4);
	\draw[color=gray, dashed] (7.5, 3.1) -- (9.1, 3.1);
	\draw[color=gray, dashed] (7.5, 3.8) -- (9.1, 3.8);
	\draw[color=gray, dashed] (7.5, 4.5) -- (9.1, 4.5);
	\draw[color=gray, dashed] (7.5, 5.4) -- (9.1, 5.4);
	\draw[color=gray, dashed] (7.5, 6) -- (9.1, 6);
	\draw[color=gray, dashed] (7.5, 6.6) -- (9.1, 6.6);
	\draw[color=gray, dashed] (7.5, 7.2) -- (9.1, 7.2);
\end{tikzpicture}
\end{tabular}
}
\caption{(A)
A realisation of the graph traversed by Hudson's algorithm started from a
sample of three chromosomes with $m$ discrete sites each at time $t = 0$, and
propagated until time $T$. The MRCA on the genetic interval $[v, w)$ is reached
at event \noderef{b}, while that on $[0, v)$ is reached at event \noderef{c}.
The non-ancestral segment $[v, w)$ above
\noderef{a} contributes to the rate of effective recombinations because it
is trapped between ancestral segments. The two columns titled CA and RE
are the respective rates of mergers and recombinations when
the recombination rate is $\rho$.
(B) A corresponding realisation of a Big ARG, which augments Hudson's algorithm
by tracking nonancestral lineages. The result is a simpler state space and
dynamics, at the cost of extra nodes and edges, highlighted in red, which do
not affect the local tree at any site.
Recombination positions are labelled alphabetically in time, and their ordering
along the genome is $y < v < x < w$, of which the first only appears in panel B.
There are two separate recombination events at link $w$.
}
\label{hudson_vs_bigARG}
\end{figure}
